Reintroduction of resistant frogs facilitates landscape-scale recovery in the presence of a lethal fungal disease

### 2024-09-27 doc/manuscript/supplementary_information.qmd
Changes made in response to editorial checklist

![**Bd loads on frogs before and after translocation.** The "before" period is immediately prior to translocation and the "after" period is the 1-year period after translocation. Bd loads are expressed as the number of ITS1 copies per skin swab, as estimated by qPCR of the Bd ITS1 region. Box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range. Samples sizes are provided immediately above the x-axis. The red dashed horizontal line indicates the approximate load above which severe disease is typically observed @vredenburg2010. This threshold (5.8 ITS1 copies - on a log10 scale) is commonly exceeded during Bd epizootics, but is exceeded only rarely in recovering populations @briggs2010.](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="80%"}

@@

![**Average yearly survival probabilities of translocated versus naturally recruited adults.** Naturally recruited individuals were the result of reproduction by translocated frogs at each site. In contrast to Fig. 3, these are not survival probabilities from the first year following translocation, but instead represent averaged survival probabilities across multiple years and cohorts. Points are median estimates and error bars give the 95% uncertainty intervals around the estimates, accounting for yearly variation in survival probabilities. All estimates were derived using the mrmr package, and the methods for calculation are described in Supplemenary Information - Population viability modeling - Estimating model parameters.](figures/compare_surv_probs.png){#fig-compare_surv_probs width="60%"}

@@

| **Parameter^a^** | **Value** | **Source** |
|:---|:---|:---|
| $\sigma_{L_1}$, Yearly survival probability of year-1 tadpoles | 0.7 | Estimated from field data, observations |
| $\sigma_{L_2}$, Yearly survival probability of year-2 tadpoles | 0.7 | Estimated from field data, observations |
| $\sigma_{L_3}$, Yearly survival probability of year-3 tadpoles | 0.7 | Estimated from field data, observations |
| $\sigma_{J_1}$, Yearly survival probability of year-1 juveniles. | Varies yearly | Varies. Bounds estimated from field data, observations |
| $\sigma_{J_2}$, Yearly survival probability of year-2 juveniles | 0.5 | Estimated from field data, observations |
| $\sigma_{A_R}$, Yearly survival probability of naturally recruited adults | Varies by population | Estimated from CMR studies |
| $\sigma_{A_T}$, Yearly survival probability of translocated adults | Varies by population | Estimated from CMR studies |
| $p_{L_1}$, Probability of a year-1 tadpoles remaining as a tadpoles | 1 | Estimated from field data, observations |
| $p_{L_2}$, Probability of a year-2 tadpoles remaining as a tadpoles | 0.25 | Estimated from field data, observations |
| $p_{J_1}$, Probability of a year-1 juvenile remaining as a juvenile | 0.25 | Estimated from field data, observations |
| $p_F$, Probability of an adult female reproducing in a year | 0.5 | Could be as high at 1, based on field observations |
| $\rho$, Proportion of adults that are female | 0.5 | Estimated from field data |
| $F$, Number of surviving eggs produced by an adult female | 100 | From observations of captive frogs |

>>>>

![**Bd loads on frogs before and after translocation.** The "before" period is immediately prior to translocation and the "after" period is the 1-year period after translocation. Bd loads are expressed as the number of ITS1 copies per skin swab, as estimated by qPCR of the Bd ITS1 region. Box plots show medians (horizontal line), first and third quartiles (hinges), largest and smallest values within 1.5x interquartile range (whiskers), and values outside the 1.5x interquartile range (dots). Sample sizes are provided immediately above the x-axis. The red dashed horizontal line indicates the approximate load above which severe disease is typically observed @vredenburg2010. This threshold (5.8 ITS1 copies - on a log10 scale) is commonly exceeded during Bd epizootics, but is exceeded only rarely in recovering populations @briggs2010. Source data are provided as a Source Data file.](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="80%"}

@@

![**Average yearly survival probabilities of translocated versus naturally recruited adults.** Naturally recruited individuals were the result of reproduction by translocated frogs at each site. In contrast to Fig. 3, these are not survival probabilities from the first year following translocation, but instead represent averaged survival probabilities across multiple years and cohorts. Points are median estimates and error bars give the 95% uncertainty intervals around the estimates, accounting for yearly variation in survival probabilities. All estimates were derived using the mrmr package, and the methods for calculation are described in Supplemenary Information - Population viability modeling - Estimating model parameters. Source data are provided as a Source Data file. ](figures/compare_surv_probs.png){#fig-compare_surv_probs width="60%"}

@@

| **Parameter^a^** | **Value** | **Source** |
|:-----------------------|:-----------------------|:-----------------------|
| $\sigma_{L_1}$, Yearly survival probability of year-1 tadpoles | 0.7 | Estimated from field data, observations |
| $\sigma_{L_2}$, Yearly survival probability of year-2 tadpoles | 0.7 | Estimated from field data, observations |
| $\sigma_{L_3}$, Yearly survival probability of year-3 tadpoles | 0.7 | Estimated from field data, observations |
| $\sigma_{J_1}$, Yearly survival probability of year-1 juveniles. | Varies yearly | Varies. Bounds estimated from field data, observations |
| $\sigma_{J_2}$, Yearly survival probability of year-2 juveniles | 0.5 | Estimated from field data, observations |
| $\sigma_{A_R}$, Yearly survival probability of naturally recruited adults | Varies by population | Estimated from CMR studies |
| $\sigma_{A_T}$, Yearly survival probability of translocated adults | Varies by population | Estimated from CMR studies |
| $p_{L_1}$, Probability of a year-1 tadpoles remaining as a tadpoles | 1 | Estimated from field data, observations |
| $p_{L_2}$, Probability of a year-2 tadpoles remaining as a tadpoles | 0.25 | Estimated from field data, observations |
| $p_{J_1}$, Probability of a year-1 juvenile remaining as a juvenile | 0.25 | Estimated from field data, observations |
| $p_F$, Probability of an adult female reproducing in a year | 0.5 | Could be as high at 1, based on field observations |
| $\rho$, Proportion of adults that are female | 0.5 | Estimated from field data |
| $F$, Number of surviving eggs produced by an adult female | 100 | From observations of captive frogs |

### 2024-09-27 doc/manuscript/translocation.qmd
Buffer equation number in main text. Auto table update
Changes made in response to editorial checklist

^6^Resources Management and Science, Yosemite National Park, El Portal, CA, 95318

@@

For the `r n_distinct(d1$site_id, d1$date)` translocations we conducted, we identified donor populations from which adult frogs ($\geq$ 40 mm snout-vent length) could be collected based on several years of VES and skin swab collections [@knapp2016], and results from population genetic analyses [@poorten2017].
The populations that we selected contained hundreds of *R. sierrae* adults and thousands of tadpoles.
These relatively high abundances were the result of recent increases following previous Bd-caused declines [@knapp2016].
As is typical for recovering *R. sierrae* populations, Bd prevalence in the donor populations was high (`r round(bdprev_min, 2)`--`r round(bdprev_max, 2)`) and Bd load (median log~10~(load) = `r round(bdload_min, 2)`--`r round(bdload_max, 2)` ITS copies) was two or more orders of magnitude below the level at which frog mortality is expected [@vredenburg2010; @joseph2018] (log~10~(load) $\approx$ 5.78 ITS copies).
Recipient sites to which frogs were translocated were chosen based on previous *R. sierrae* presence (determined from VES and/or museum records) or characteristics that suggested high quality habitat for this species [@knapp2005].
At the beginning of this study, we had a relatively limited understanding of the factors that affect habitat quality.
In subsequent years, we improved our site selection process by incorporating new information about important habitat features, in particular, overwintering habitats such as submerged boulders and overhanging banks.
*R. sierrae* were absent from all recipient sites prior to the first translocation.
This determination was based on VES conducted at recipient sites over several years (see additional details below regarding frog detectability during VES).

@@

During each site visit, we conducted VES either immediately before CMR surveys or during the first day of CMR surveys.
VES was conducted by walking the entire water body perimeter, first 100 m of each inlet and outlet stream, and any fringing ponds and wetlands, and counting all *R. sierrae* tadpoles and juveniles.
These *R. sierrae* life stages have high detectability, and counts are highly repeatable and provide estimates of relative abundance [@knapp2000].
The above-described collection of data from live *R. sierrae* was approved under the University of California-Santa Barbara IACUC protocol #478.

@@

$$
\begin{bmatrix}
L_1 \\
L_2 \\
L_3 \\
J_1 \\
J_2 \\
A_R \\ 
A_T
\end{bmatrix}(t + 1) = 
\begin{bmatrix}
  0 & 0 & 0 & 0 & 0 & \sigma_{A_R} p_F \rho F & \sigma_{A_T} p_F \rho F \\
  \sigma_{L1} p_{L1} & 0 & 0 & 0 & 0 & 0 & 0 \\
  0 & \sigma_{L2} p_{L2} & 0 & 0 & 0 & 0 & 0\\
  \sigma_{L1} (1 - p_{L1}) & \sigma_{L2} (1 - p_{L2}) & \sigma_{L3} & 0 & 0 & 0 & 0 \\
  0 & 0 & 0 & \sigma_{J1} p_{J1} & 0 & 0 & 0 \\
  0 & 0 & 0 & \sigma_{J1} (1 - p_{J1}) & \sigma_{J2}  & \sigma_{A_R} & 0 \\
  0 & 0 & 0 & 0 & 0 & 0 & \sigma_{A_T} 
\end{bmatrix} \begin{bmatrix}
L_1 \\
L_2 \\
L_3 \\
J_1 \\
J_2 \\
A_R \\
A_T \\
\end{bmatrix}(t) 
$$

@@

The data used in this study are available at <https://github.com/SNARL1/translocation>.
The CMR datasets from each reintroduction site are available at <https://github.com/SNARL1/cmr-analysis>.

@@

R code to replicate our analyses is available at <https://github.com/SNARL1/translocation>.
R code to analyze the CMR datasets is available at <https://github.com/SNARL1/cmr-analysis>.

@@

We thank the following for important contributions to this study: E.
Hegeman and A.
Lindauer (project and data management, field and laboratory work); A.
Barbella and K.
Rose (laboratory work); numerous summer technicians (field work); staff at Sequoia-Kings Canyon and Yosemite National Parks, Inyo and Sierra National Forests, California Department of Fish and Wildlife, U.S. Fish and Wildlife Service, University of California-Santa Barbara Institutional Animal Care and Use Committee, and Sierra Nevada Aquatic Research Laboratory (research permits, logistical support, and/or field assistance).
This project was supported by grants from the National Park Service (to R.A.K.), Yosemite Conservancy (to R.A.K.), and National Science Foundation (EF-0723563, to C. Briggs; DEB-1557190, to C. Briggs; DEB-2133401, to M.Q.W; and DBI-2120084, to C. Richards-Zawacki).

@@

![**Conditional effects of predictors of 1-year frog survival**. Results are from the among-site meta-analysis and 1-year frog survival is expressed as a probability. **a** winter severity in the year following translocation, **b** elevation of recipient site, **c** donor population, and **d** sex. In (a) and (b), blue lines are medians and gray ribbons are 95% uncertainty intervals. In (c) and (d), box plots show medians, first and third quartiles, largest and smallest values within 1.5x interquartile range, and values outside the 1.5x interquartile range.](figures/cond_effects_plot.png){#fig-cond-effects width="80%"}

>>>>

^6^Resources Management and Science Division, Yosemite National Park, El Portal, CA, 95318

@@

The collection of data from live *R. sierrae* described here complies with all ethical regulations and was approved under the University of California-Santa Barbara Institutional Animal Care and Use Committe (IACUC) protocol number 478.
For the `r n_distinct(d1$site_id, d1$date)` translocations we conducted, we identified donor populations from which adult frogs ($\geq$ 40 mm snout-vent length) could be collected based on several years of VES and skin swab collections [@knapp2016], and results from population genetic analyses [@poorten2017].
The populations that we selected contained hundreds of *R. sierrae* adults and thousands of tadpoles.
These relatively high abundances were the result of recent increases following previous Bd-caused declines [@knapp2016].
As is typical for recovering *R. sierrae* populations, Bd prevalence in the donor populations was high (`r round(bdprev_min, 2)`--`r round(bdprev_max, 2)`) and Bd load (median log~10~(load) = `r round(bdload_min, 2)`--`r round(bdload_max, 2)` ITS copies) was two or more orders of magnitude below the level at which frog mortality is expected [@vredenburg2010; @joseph2018] (log~10~(load) $\approx$ 5.78 ITS copies).
Recipient sites to which frogs were translocated were chosen based on previous *R. sierrae* presence (determined from VES and/or museum records) or characteristics that suggested high quality habitat for this species [@knapp2005].
At the beginning of this study, we had a relatively limited understanding of the factors that affect habitat quality.
In subsequent years, we improved our site selection process by incorporating new information about important habitat features, in particular, overwintering habitats such as submerged boulders and overhanging banks.
*R. sierrae* were absent from all recipient sites prior to the first translocation.
This determination was based on VES conducted at recipient sites over several years (see additional details below regarding frog detectability during VES).

@@

During each site visit, we conducted VES either immediately before CMR surveys or during the first day of CMR surveys.
VES was conducted by walking the entire water body perimeter, first 100 m of each inlet and outlet stream, and any fringing ponds and wetlands, and counting all *R. sierrae* tadpoles and juveniles.
These *R. sierrae* life stages have high detectability, and counts are highly repeatable and provide estimates of relative abundance [@knapp2000].

@@

$$
\begin{bmatrix}
L_1 \\
L_2 \\
L_3 \\
J_1 \\
J_2 \\
A_R \\ 
A_T
\end{bmatrix}(t + 1) = 
\begin{bmatrix}
  0 & 0 & 0 & 0 & 0 & \sigma_{A_R} p_F \rho F & \sigma_{A_T} p_F \rho F \\
  \sigma_{L1} p_{L1} & 0 & 0 & 0 & 0 & 0 & 0 \\
  0 & \sigma_{L2} p_{L2} & 0 & 0 & 0 & 0 & 0\\
  \sigma_{L1} (1 - p_{L1}) & \sigma_{L2} (1 - p_{L2}) & \sigma_{L3} & 0 & 0 & 0 & 0 \\
  0 & 0 & 0 & \sigma_{J1} p_{J1} & 0 & 0 & 0 \\
  0 & 0 & 0 & \sigma_{J1} (1 - p_{J1}) & \sigma_{J2}  & \sigma_{A_R} & 0 \\
  0 & 0 & 0 & 0 & 0 & 0 & \sigma_{A_T} 
\end{bmatrix} \begin{bmatrix}
L_1 \\
L_2 \\
L_3 \\
J_1 \\
J_2 \\
A_R \\
A_T \\
\end{bmatrix}(t)
$$ {#eq-matrix}

@@

The data used in this study are available at <https://github.com/SNARL1/translocation> [@knapp2024a].
This includes the estimated survival of each reintroduced frog cohort, data used to identify predictors of 1-year frog survival, data used to estimate growth rates and extinction probabilities of reintroduced populations, Bd loads on frogs before and after translocation, and survival probabilities of translocated versus naturally recruited adults.
For the relevant figures, source data are provided in a Source Data file.
The CMR datasets from each reintroduction site that we used to calculate post-translocation frog survival are available at <https://github.com/SNARL1/cmr-analysis> [@knapp2024b].

@@

R code to replicate our analyses is available at <https://github.com/SNARL1/translocation> [@knapp2024a].
R code to analyze the CMR datasets is available at <https://github.com/SNARL1/cmr-analysis> [@knapp2024b].

@@

We thank the following for important contributions to this study: E.
Hegeman and A.
Lindauer (project and data management, field and laboratory work); A.
Barbella and K.
Rose (laboratory work); numerous summer technicians (field work); staff at Sequoia-Kings Canyon and Yosemite National Parks, Inyo and Sierra National Forests, California Department of Fish and Wildlife, U.S. Fish and Wildlife Service, Sierra Nevada Aquatic Research Laboratory (DOI: 10.21973/N3966F), and University of California-Santa Barbara Institutional Animal Care and Use Committee (research permits, logistical support, and/or field assistance).
This project was supported by grants from the National Park Service (to R.A.K.), Yosemite Conservancy (to R.A.K.), and National Science Foundation (EF-0723563, to C. Briggs; DEB-1557190, to C. Briggs; DEB-2133401, to M.Q.W; and DBI-2120084, to C. Richards-Zawacki).

@@

![**Conditional effects of predictors of 1-year frog survival**. Results are from the among-site meta-analysis and 1-year frog survival is expressed as a probability. **a** winter severity in the year following translocation, **b** elevation of recipient site, **c** donor population, and **d** sex. In (a) and (b), blue lines are medians and gray ribbons are 95% uncertainty intervals for the posterior distributions of the predicted effects. In (c) and (d), box plots show medians (horizontal line), first and third quartiles (hinges), largest and smallest values within 1.5x interquartile range (whiskers), and values outside the 1.5x interquartile range (dots) for the posterior distributions of the predicted effects.](figures/cond_effects_plot.png){#fig-cond-effects width="80%"}
